---
title: "Lab_8"
author: "Eleanor Lindsey"
format: html
editor: visual
execute:
  echo: true
---

## 

install.packages('visdat')

```{r}
library(dplyr)
library(tidyverse)
library(tidymodels)
library(powerjoin)
library(glue)
library(vip)
library(baguette)
library(visdat)
library(ggpubr)
library(ggplot2)
```

### Load in the data

```{r}
root  <- 'https://gdex.ucar.edu/dataset/camels/file'
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_attributes_v2.0.pdf', 
              'data/camels_attributes_v2.0.pdf')
types <- c("clim", "geol", "soil", "topo", "vege", "hydro")
# Where the files live online ...
remote_files  <- glue('{root}/camels_{types}.txt')
# where we want to download the data ...
local_files   <- glue('data/camels_{types}.txt')
walk2(remote_files, local_files, download.file, quiet = TRUE)
# Read and merge data
camels <- map(local_files, read_delim, show_col_types = FALSE) 
camels <- power_full_join(camels ,by = 'gauge_id')
```

### Clean the Data

```{r}
camels%>%
  select(aridity, p_mean, q_mean)%>%
  drop_na()%>% 
  cor()

plot_1<-ggplot(data = camels, aes(x = gauge_lon, y = gauge_lat)) +
  borders("state", colour = "gray50") +
  geom_point(aes(color = aridity)) +
  labs(title='Levels of Aridity',
       x='Longitude',
       y='Latitude') +
  scale_color_gradient(low = "dodgerblue", high = "forestgreen") +
  ggthemes::theme_map()

plot_2 <- ggplot(data = camels, aes(x = gauge_lon, y = gauge_lat)) +
  borders("state", colour = "gray50") +
  geom_point(aes(color = p_mean)) +
  labs(title = 'Mean Daily Precipitation',
       x = 'Longitude',
       y = 'Latitude') +
  scale_color_gradient(low = "purple", high = "pink") +
  ggthemes::theme_map()

ggarrange(plot_1, plot_2, n_col=2)

```

### Visual EDA

```{r}
# Create a scatter plot of aridity vs rainfall
ggplot(camels, aes(x = aridity, y = p_mean)) +
  # Add points colored by mean flow
  geom_point(aes(color = q_mean)) +
  # Add a linear regression line
  geom_smooth(method = "lm", color = "red", linetype = 2) +
  # Apply the viridis color scale
  scale_color_viridis_c() +
  # Add a title, axis labels, and theme (w/ legend on the bottom)
  theme_linedraw() + 
  theme(legend.position = "bottom") + 
  labs(title = "Aridity vs Rainfall vs Runnoff", 
       x = "Aridity", 
       y = "Rainfall",
       color = "Mean Flow")
```

```{r}
ggplot(camels, aes(x = aridity, y = p_mean)) +
  geom_point(aes(color = q_mean)) +
  geom_smooth(method = "lm") +
  scale_color_viridis_c() +
  # Apply log transformations to the x and y axes
  scale_x_log10() + 
  scale_y_log10() +
  theme_linedraw() +
  theme(legend.position = "bottom") + 
  labs(title = "Aridity vs Rainfall vs Runnoff", 
       x = "Aridity", 
       y = "Rainfall",
       color = "Mean Flow")
```

```{r}
ggplot(camels, aes(x = aridity, y = p_mean)) +
  geom_point(aes(color = q_mean)) +
  geom_smooth(method = "lm") +
  # Apply a log transformation to the color scale
  scale_color_viridis_c(trans = "log") +
  scale_x_log10() + 
  scale_y_log10() +
  theme_linedraw() +
  theme(legend.position = "bottom",
        # Expand the legend width ...
        legend.key.width = unit(2.5, "cm"),
        legend.key.height = unit(.5, "cm")) + 
  labs(title = "Aridity vs Rainfall vs Runnoff", 
       x = "Aridity", 
       y = "Rainfall",
       color = "Mean Flow") 
```

### Splitting the Data

```{r}
set.seed(123)
# Bad form to perform simple transformations on the outcome variable within a 
# recipe. So, we'll do it here.
camels <- camels |> 
  mutate(logQmean = log(q_mean))

# Generate the split
camels_split <- initial_split(camels, prop = 0.8)
camels_train <- training(camels_split)
camels_test  <- testing(camels_split)

camels_cv <- vfold_cv(camels_train, v = 10)
```

### Creating a Recipe

```{r}
rec <-  recipe(logQmean ~ aridity + p_mean, data = camels_train) %>%
  # Log transform the predictor variables (aridity and p_mean)
  step_log(all_predictors()) %>%
  # Add an interaction term between aridity and p_mean
  step_interact(terms = ~ aridity:p_mean) |> 
  # Drop any rows with missing values in the pred
  step_naomit(all_predictors(), all_outcomes())
```

### Defining Models and Workflows

```{r}
# Define model
lm_model <- linear_reg() %>%
  # define the engine
  set_engine("lm") %>%
  # define the mode
  set_mode("regression")

# Instantiate a workflow ...
lm_wf <- workflow() %>%
  # Add the recipe
  add_recipe(rec) %>%
  # Add the model
  add_model(lm_model) %>%
  # Fit the model to the training data
  fit(data = camels_train) 

rf_model <- rand_forest() %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("regression")

rf_wf <- workflow() %>%
  # Add the recipe
  add_recipe(rec) %>%
  # Add the model
  add_model(rf_model) %>%
  # Fit the model
  fit(data = camels_train) 

b_model<- boost_tree()%>%
  set_engine('xgboost')%>%
  set_mode('regression')

bm_wf<- workflow()%>%
  add_recipe(rec)%>%
  add_model(b_model)%>%
  fit(data=camels_train)
```

### Predictions

```{r}
lm_data <- augment(lm_wf, new_data = camels_test)
dim(lm_data)

rf_data <- augment(rf_wf, new_data = camels_test)
dim(rf_data)

bm_data<- augment(bm_wf, new_data=camels_test)
dim(bm_data)
```

### Model Evaluations

```{r}
metrics(lm_data, truth = logQmean, estimate = .pred)

ggplot(lm_data, aes(x = logQmean, y = .pred, colour = aridity)) +
  scale_color_viridis_c() +
  geom_point() +
  geom_abline() +
  labs(title='Linear Regression Model Evaluation')+
  theme_linedraw()

metrics(rf_data, truth = logQmean, estimate = .pred)

ggplot(rf_data, aes(x = logQmean, y = .pred, colour = aridity)) +
  scale_color_viridis_c() +
  geom_point() +
  geom_abline() +
  labs(title='Random Forest Model Evaluation')+
  theme_linedraw()

metrics(bm_data, truth = logQmean, estimate = .pred)

ggplot(bm_data, aes(x = logQmean, y = .pred, colour = aridity)) +
  scale_color_viridis_c() +
  geom_point() +
  geom_abline() +
  labs(title='BoostTree Model Evaluation')+
  theme_linedraw()
```

### Workflow Approach

```{r}
wf <- workflow_set(list(rec), list(Linear=lm_model, Ranger=rf_model, Boost=b_model)) %>%
  workflow_map('fit_resamples', resamples = camels_cv)%>%
  workflow_map(resamples=camels_cv, 
               metrics=metric_set(rsq, rmse, mae))

autoplot(wf)

rank_results(wf, rank_metric = "rsq", select_best = TRUE)
```

Based on the testing of the random forest, boost tree, and linear regression models, the random forest model is the best choice. It has a low mean absolute error and rmse which mean there are low levels of errors detected.

The model I selected is the Random forest model with a ranger engine and a regression mode. It works well because it is designed to take multiple predictors and return a prediction based on hyperparameters within the data.

### Building the Specified Model

```{r}

# Define the model with tuneable parameters
rf_model <- rand_forest(min_n = tune(), trees = tune()) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("regression")

# 2. Create the workflow
rf_wf_tuned <- workflow() %>%
  add_recipe(rec) %>%
  add_model(rf_model)
```

### Check the Tuneable Values

```{r}
dials<-extract_parameter_set_dials(rf_wf_tuned)

dials$object
```

### Defining the Search Space

```{r}
my.grid<-dials%>%
  grid_latin_hypercube(size = 20)
```

### Tuning the Model

```{r}
model_params <-  tune_grid(
    rf_wf_tuned,
    resamples = camels_cv,
    grid = my.grid,
    metrics = metric_set(rmse, rsq, mae),
    control = control_grid(save_pred = TRUE)
  )

autoplot(model_params)
```

From what I see as minimal node size decreases, so does rmse and mae while rsq increases. This indicates high correlation and low error between the data set and the minimal node size. There is no obvious correlation between the number of trees and the data set.

### Check the Skill of the Tuned Model

```{r}
collect_metrics(model_params)
tree_metrics = metric_set(rsq, rmse, mae)

```

#### Collect Metrics Interpretation

From the table I can interpret that data has a very low standard error which supports the significance of the testing.

#### Show Best

```{r}
show_best(model_params, metric = "rsq") 
show_best(model_params, metric = "rmse") 
show_best(model_params, metric = "mae")
```

#### Show Best Interpretation

based on MAE the best hyper parameter for this test is the set with 336 trees and a 25 minimum number of trees

```{r}
hp_best<- show_best(model_params, metric = "mae", n = 1)

print(hp_best)
```

#### Finalizing the Model

```{r}
final_wf<-finalize_workflow(rf_wf_tuned,hp_best)

```

### Final Model Verification

```{r}
final_fit <- last_fit(final_wf, camels_split, metrics =tree_metrics )

collect_metrics(final_fit)

```

#### Collect Metrics Interpretation

The r squared value (rsq) explains about 75% of the variance in the random forest model. This value demonstrates relatively strong statistical evidence in the correlation of the variables.

An RMSE of 0.5678406 indicates that, on average, the model's predictions are off by about 0.568 units in the original scale of the target variable.

An MAE of 0.3497950 means that, on average, the model’s predictions are off by about 0.350 units. The lower value indicates strong preformance.

#### Collect Predictions

```{r}
collect_predictions(final_fit) |> 
  ggplot(aes(x = .pred, y = logQmean)) + 
  geom_point() +
  geom_abline() + 
  geom_smooth(method = "lm") + 
  theme_linedraw() + 
  labs(title = "Final Fit", 
       x = "Predicted (Log10)", 
       y = "Actual (Log10)")
```

### Building a Map

#### Augment

```{r}
full_pred = fit(final_wf, data = camels) |>
  augment(new_data = camels) 

```

#### Mutate

```{r}
full_pred_mutated<- full_pred%>%
  mutate(residuals=(.pred-logQmean)^2)
colnames(full_pred_mutated)
```

#### ggplot Predictions

```{r}
predictions<-ggplot(full_pred, aes(x=logQmean, y=.pred))+
  geom_point()+
  geom_smooth(method = 'lm')+
  labs(x="LogQMean", y="predictions", title="Random Forest Model Predictions")
predictions
```

#### ggplot Residuals

```{r}
residuals<-ggplot(full_pred_mutated, aes(x=logQmean, y=residuals))+
  geom_point()+
  geom_smooth(method='lm')+
  labs(x="LogQMean", y="residuals", title="Residuals for Random Forest Model")
residuals
```

#### Combined Figure

```{r}
ggarrange(residuals, predictions)
```
